### 2014-08-07 writeups/context-dependence.tex
hutch talk done
A few comments and fixes.

  \begin{center}
    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
      CG   &  TG   &  $\gamma$ & \\
      CG   &  CA   &  $\gamma$ &
    \end{tabular} .
  \end{center}
  Here $\gamma$ is the additional CpG rate above the base mutation rate,
  and a total of 13 parameters.
  Note that this model is very different in spirit to TASEP,
  in which the total number of particles was conserved.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation''.
Briefly, this \todo{This is called the Gillespie algorithm, right?} works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\todo{A little confused here. Previously, the $j$ in $u^j$ was used to index the possible patterns, right? Now it's indexing possible changes at a site.}
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@

\todo[inline]{In this section I believe that we are talking about the full likelihood, rather than the likelihood of just non-overlapping windows of the data.
I'm also a little confused about "choosing" a subset of patterns.
These pattern choices are determining our model?}

>>>>

  \begin{center}
    \begin{tabular}{c@{\quad$\to$\quad}c@{\quad at rate\quad }cc}
      $x$  &  $y$  &  $m_{xy}$ & when $x \neq y \in \{\nA,\nC,\nG,\nT\}$  \\
      \nC\nG   &  \nT\nG   &  $\gamma$ & \\
      \nC\nG   &  \nC\nA   &  $\gamma$ &
    \end{tabular} .
  \end{center}
  Here $\gamma$ is the additional CpG rate above the base mutation rate,
  and a total of 13 parameters.
  Note that this model is very different in spirit to TASEP,
  in which the total number of particles was conserved.

@@

First, it will be useful for later proofs to have an explicit construction of the process,
also used for simulation,
using a variant of the ``jump chain representation'', also known as the ``Gillespie algorithm.''
.
Briefly, this  works by first sampling a homogeneous Poisson process of times and locations of possible changes,
at an appropriate rate $\mu_*$ per site,
and then resolving each possible change in temporal order.
The rate $\mu_*$ should be the maximum ``local'' rate at which transitions occur, across sites and transition outcome states:
\[
\mu_* = \max \left\{ \sum_{y \in \calS^n} \sum_{j\in J(i,x,y)} \mu_j \qquad \st x \in \calS^n, \quad 1 \le i \le n \right\} .
\]
Note that since mutation is homogeneous, the maximum rate of changes occurring at any position will be the same, except possibly near the end of the sequence,
so the maximization over $i$ is unnecessary.
Now sample possible changes: let $c_{ij}$ denote the time of the $j^\text{th}$ possible change at site $i$,
so that for each site $i$,
\[
0 \le c_{i1} < c_{i2} < \cdots < c_{iN_i} \le t
\]
is a draw from a rate $\mu_*$ Poisson process on $[0,t]$.
These processes are independent across sites.
Now, suppose we have determined the state at time $s$ to be $X(s) = x$,
and the next event is $c_{ij} = \min_{k,h} \{ c_{kh} : c_{kh}>s \}$.
At $c_{ij}$,
a new state is then chosen at position $i$
with probability proportional to the corresponding rate, i.e.\
\todo{A little confused here. Previously, the $j$ in $u^j$ was used to index the possible patterns, right? Now it's indexing possible changes at a site.}
\[
q_j = \begin{cases}
  \mu_j/\mu_* \qquad & \text{if } x_i^{(|u^j|)} = u^j  \\
  0 \qquad & \text{otherwise,}
\end{cases}
\]
is the probability that $x_i^{(|u^j|)}$ is replaced with $v^j$.
The remaining probability $q_0 = 1-\sum_{j=1}^{|\calT|} q_j$ gives the probability that the state remains the same.
(By construction of $\mu_*$, $q_0\ge 0$.)

@@
